An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-07 main.tex
skip duels

Parikh images are also applicable to decide more general classes of
string constraints~\cite{ostrich-plus}. Consider, for instance, a
constraint involving the substring operation:
\begin{equation}
  \label{eq:substring}
  x \in \Language_1 \wedge \; 0 \leq n \leq m < |x| \wedge x[n:m] \in \Language_2
\end{equation}
in which $x$ is a string variable, $n, m$ are integer variables, and
$x[n:m]$ denotes the substring of $x$ starting at position~$n$ running
until position~$m$. To illustrate the decision procedure in
\cite{ostrich-plus} for string constraints of this kind, suppose that
$\Language_2$ is defined by the regular
expression~$\mathtt{(ab|c)*}$. An automaton recognising this language
is shown in Fig.~\ref{fig:parikh-automata} (left).

We can model \eqref{eq:substring} using the notion of a \emph{Parikh
  automaton}~\cite{parikh-automata,expressiveness}. A Parikh automaton
is an automaton in which transitions are labelled both with characters
from the alphabet~$\Alphabet$, and with offset vectors defining the
increments of a finite number of counters.

In the following definitions, we use (as in~\cite{expressiveness}) the following notation:
\begin{itemize}
  \item $\Alphabet$ is an alphabet
  \item $D$ is a finite subset of $\Naturals^d$, for some $d \in \Naturals^+$ (e.g. vectors of length $d$)
\end{itemize}

\begin{definition}\label{def:projections}
  We define two projections on the composite alphabet $\left(\Alphabet \times D\right)^*$:
  \begin{itemize}
    \item $\pi_{\Alphabet} := (a, \Vector{v}) \mapsto a$ or the projection on $\Alphabet$ is the monoid morphism from $\left(\Alphabet \times D\right)^*$ to $\Alphabet^*$, and the opposite projection
    \item $\pi_{v} := (a, \Vector{v}) \mapsto \Vector{v}$, called the
    \emph{extended Parikh image}.
  \end{itemize}
\end{definition}

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where $C \subseteq \Naturals^d$ is
  a semi-linear set (or, equivalently, a Presburger formula), and $\Automaton$
  is a finite automaton with the alphabet $\Alphabet \times D$, where $D
  \subseteq \Naturals^d$. We say that $\Tuple{\Automaton, \varphi}$
  \emph{recognises} the language, $\Language(\Automaton, C)$, that is it
  recognises the language created by is the $\pi_\Alphabet$-projection on
  $\Alphabet$ whose $\pi_{v} \subseteq C$.
\end{definition}

@@

\begin{figure}[t]
  \centering
  \includegraphics[scale=0.9]{counter_automaton_pre}
  \hfill
  \includegraphics[scale=0.9]{counter_automaton}
  \caption{An automaton recognising the language $\mathtt{(ab|c)*}$
    (left), and a Parikh automaton describing the pre-image of the
    substring operation (right).}\label{fig:parikh-automata}
    \end{figure}

An example of a Parikh automata version of $\AcaOrBc{}$ can be seen in
\cref{fig:parikh-automaton}. A corresponding Presburger version of the
constraint seen in \cref{sec:motivation} is $\psi(\Vec{v}) = v_1 > v_2$.

Generally, we can encode the problem as follows:
\begin{enumerate}
  \item The morphism ($\Map$) we would use is $\pi_{v}$
  \item The monoid operation is element-wise vector addition, which is trivially commutative.
  \item \Fudge{We would just add the constraints on the final values!?}
\end{enumerate}

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.75\linewidth}
    \centering 
      \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter}
      \caption{Runtime duel: \Calculus{} versus baseline.}
      \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter}
      \caption{Runtime duel: \Calculus{} versus \Nuxmv.}
    \end{minipage}  
  \caption{The pairwise runtime of each instance for instances solved by any backend. Timeouts are reported as a runtime of \RuntimeTimeout.}
  \label{fig:duels}
\end{figure}

>>>>

Parikh images are also applicable to decide more general classes of
string constraints~\cite{ostrich-plus}. Consider, for instance, a
constraint involving the substring operation:
\begin{equation}
  \label{eq:substring}
  x \in \Language_1 \wedge \; 0 \leq n \leq m \leq |x| \wedge x[n:m] \in \Language_2
\end{equation}
in which $x$ is a string variable, $n, m$ are integer variables, and
$x[n:m]$ denotes the substring of $x$ starting at position~$n$ running
until position~$m$. This constraint belongs to an expressive fragment
of string logic that cannot be decided by most state-of-the-art string
solvers. To illustrate the decision procedure proposed in
\cite{ostrich-plus} for string constraints of this kind, suppose that
$\Language_2$ is defined by the regular
expression~$\mathtt{(ab|c)*}$. An automaton recognising this language
is shown in Fig.~\ref{fig:parikh-automata} (left).

We can model \eqref{eq:substring} using the notion of a \emph{Parikh
  automaton}~\cite{parikh-automata,expressiveness}. A Parikh automaton
is an automaton in which transitions are labelled both with characters
from the alphabet~$\Alphabet$, and with offset vectors defining the
increments of a finite number of counters. This means that Parikh
automata recognise words over an extended
alphabet~$\Alphabet \times D$, where $D \subseteq \Naturals^d$ is a
finite set of increment vectors (notation as
in~\cite{expressiveness}). We use the symbols~$\pi_\Alphabet, \pi_D$
to denote projections to the first and the second component of a
composite letter~$(a, \Vector{v})$, respectively, and extend those
projections to words:
\begin{equation*}
  \pi_\Alphabet((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ a_1 \RepeatSum{\Concat} a_k,
  \qquad
  \pi_D((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ \Vector{v}_1 \RepeatSum{+} \Vector{v}_k.
\end{equation*}

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where
  $C \subseteq \Naturals^d$ is a semi-linear set (or, equivalently, a
  Presburger formula), and $\Automaton$ is a finite automaton with the
  alphabet $\Alphabet \times D$, where $D \subseteq \Naturals^d$. We
  say that $\Tuple{\Automaton, \varphi}$ recognises a
  word~$w \in \Alphabet^*$ if and only if the automaton has a run
  accepting an extended word~$w' \in (\Alphabet \times D)^*$ such that
  $\pi_\Alphabet(w') = w$ and $\pi_D(w') \in C$.
\end{definition}

The decision procedure in \cite{ostrich-plus} for formulas like
\eqref{eq:substring} will construct a pre-image of $\Language_2$ under
the substring operation~$\cdot[n:m]$, and check whether this pre-image
is consistent with the constraint~$x \in \Language_1$. Because the
substring operation depends on the values of the integer
variables~$n, m$, it is appropriate to describe the pre-image as
Parikh automaton of dimension~$2$, resulting in the automaton in
Fig.~\ref{fig:parikh-automata} (right). The first component of the
increment vectors is used to count the number of letters eliminated in
the beginning of the string (value~$n$), while the second component
records the beginning of the eliminated suffix (value~$m$).

\Fudge{continue}

@@

\begin{figure}[t]
  \centering
  \raisebox{5ex}{\includegraphics[scale=0.9]{counter_automaton_pre}}
  \hfill
  \includegraphics[scale=0.9]{counter_automaton}
  \caption{An automaton recognising the language $\mathtt{(ab|c)*}$
    (left), and a Parikh automaton describing the pre-image of this
    language under the substring operation~$\cdot[n:m]$ (right). For
    sake of presentation, the Parikh automaton contains
    $\epsilon$-transitions, which could be eliminated in the standard
    way.}\label{fig:parikh-automata}
    \end{figure}

\iffalse
An example of a Parikh automata version of $\AcaOrBc{}$ can be seen in
\cref{fig:parikh-automaton}. A corresponding Presburger version of the
constraint seen in \cref{sec:motivation} is $\psi(\Vec{v}) = v_1 > v_2$.

Generally, we can encode the problem as follows:
\begin{enumerate}
  \item The morphism ($\Map$) we would use is $\pi_{v}$
  \item The monoid operation is element-wise vector addition, which is trivially commutative.
  \item \Fudge{We would just add the constraints on the final values!?}
\end{enumerate}
\fi

@@

% \begin{figure}[ht]
%   \begin{minipage}[b]{0.75\linewidth}
%     \centering 
%       \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter}
%       \caption{Runtime duel: \Calculus{} versus baseline.}
%       \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter}
%       \caption{Runtime duel: \Calculus{} versus \Nuxmv.}
%     \end{minipage}  
%   \caption{The pairwise runtime of each instance for instances solved by any backend. Timeouts are reported as a runtime of \RuntimeTimeout.}
%   \label{fig:duels}
% \end{figure}
